---
title: "Geodetic_pt_compare"
author: "G. Perkins"
date: "18/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE)

```

# Point comparison with geodetic mascot points

Compare the bc Data catalogue historic high precision [point data](https://catalogue.data.gov.bc.ca/dataset/mascot-geodetic-control-monuments-high-precision) with ensemble DTM, TRIM and LiDAR where possible to assess variability. 

Note a copy of the data was obtained from the bc catalogue and saved on the sync drive

```{r load data, echo = FALSE }

# Read in geometic data set (independent of data assembly process) and compare to 
# ensemble dtm, lidar, trim. 

library(raster)
library(sf)
library(bcdata)
library(bcmaps)
library(dplyr)
library(rvest)
library(ggplot2)
library(tidyr)


## Load in eDTM layer and trim (from sync location)

data.dir <- "D:/Hengle_BCDEM30"
sync.dir <- "C:/Sync/BC_DEM"
temp.dir <- file.path(data.dir,"temp")

geopt <- st_read(file.path(sync.dir, 'Point_sources', "geodetic_control", "Mascot_GeoDetic_Ctrl.shp"), quiet = TRUE)


if(file.exists(file.path(temp.dir, "mascot_dem_pts.rds"))){
  
  #print("loading file")
  geodf <- readRDS(file.path(temp.dir, "mascot_dem_pts.rds"))

} else {

  #print("generating file, be patient")
  
edtm <- raster(file.path(data.dir, "BC30m","dtm_elev.lowestmode_gedi.eml_m_30m_0..0cm_2000..2018_bc.epsg3005_v0.1.tif"))
trim <- raster(file.path(data.dir, "trim", "TRIM","bc_elevation_25m_bcalb.tif"))

geopt <- geopt %>%
  select(GEODETIC_C, LATITUDE, LONGITUDE, MUNICIPALI, QUALITY_CL, 
         LAST_UPDAT, ELEVATION,VERTICAL_D, VERTICAL_S,
         GEOID_UNDU) %>%
  mutate(edtm = raster::extract(edtm, geopt)) %>%
  mutate(trim = raster::extract(trim, geopt))

# create delta 
geodf <- geopt %>%
  mutate(dedtm = round(ELEVATION,0) - edtm,
         dtrim = round(ELEVATION,0) - trim) %>%
  filter(!is.na(dedtm))
  
saveRDS(geodf, file = file.path(temp.dir, "mascot_dem_pts.rds"))

st_write(geodf, file.path(temp.dir, "mascot_pt_att.gpkg"))

} 


```

The delta was calculated for each data type by subtracting from the elevaltion of the geodetic points. ie elevation - trim. Positive values indicate the dtm is lower than geo points, while negative points indicate the dtm is higher than the geopts. 


## Compare geopts to ensemble DTM

As point quality data is included in the mascot dataset I checked quality of points for trends in difference 

```{r clean data, echo = FALSE}
# clean and extract dataset

delta_pts <- ggplot(geodf, aes(dedtm))+ 
  geom_histogram(binwidth = 10) 

delta_pts

```

* most points centred around zero or positive side or zero
* Range in values of `r range(geodf$dedtm)`


As the data quality of points documented, I also compared delta by quality type; A - high precision, B; good, C: Poor, E; unknown quality (could be A-E).

```{r}
delta_quality <- ggplot(geodf, aes(dedtm))+ 
  geom_histogram(binwidth = 10) +
  facet_wrap(~QUALITY_CL, scales = "free_x")

delta_quality


edtm_elev <- ggplot(geodf, aes(dedtm, ELEVATION))+ 
  geom_point()+ 
  geom_vline(xintercept = 0, colour = "red")

edtm_elev

```

Also check the variability across municipalities. 

```{r}
delta_quality1 <- ggplot(geodf, aes(dedtm, MUNICIPALI))+ 
  geom_point() + 
  geom_vline(xintercept = 0, colour = "red")
delta_quality1
```

* Spread largly even across municipalitied, 
* Highly variable across rural municipalities 
* exceptions such as Prince George, 


## Compare the geopts with TRIM 

The range in difference between the mascot geodetic pts and TRIM was `r range(geodf$dtrim, na.rm= T)`.

```{r}
trim_quality <- ggplot(geodf, aes(dtrim))+ 
  geom_histogram(binwidth = 10) +
  facet_wrap(~QUALITY_CL, scales = "free_x")

trim_quality


trim_elev <- ggplot(geodf, aes(dtrim, ELEVATION))+ 
  geom_point()+ 
  geom_vline(xintercept = 0, colour = "red")

trim_elev

trim_munic <- ggplot(geodf, aes(dtrim, MUNICIPALI))+ 
  geom_point()+ 
  geom_vline(xintercept = 0, colour = "red")

trim_munic 
```
* higher uncertainty around areas with lower elevation in general 


# Delta Ensemble DTM vs Delta TRIM
```{r}

geo_long <- geodf %>%
  pivot_longer(cols = c(dedtm, dtrim), names_to = "dem_type", values_to = "value") %>%
  dplyr::select(ELEVATION, dem_type, value, geometry) 

geo_mean <-  geo_long %>%
  group_by(dem_type) %>%
  summarise(mean = mean(value, na.rm = TRUE), 
            sd = sd(value, na.rm = TRUE))

geo_mean 


delta_pt <- ggplot(geodf, aes(x = dedtm, y = dtrim))+ 
  geom_point()

delta_pt

```

```{r}

geo_long <- geo_long %>%
  mutate(
    group = case_when(
      value < -100 ~ "extremely higher (-100)", 
      value >-100 & value < -10 ~ "moderately higher (-100 to -10)", 
      value >=-10 & value < 10 ~ "correct within 10m",
    value >=10 & value <25 ~ "slight lower (10 to 25)",
    value >=25 & value <50 ~ "moderate lower (25 to 50)",
    value >=50 & value <100 ~ "large lower (50 to 100)",
    value >=100 & value <800 ~ "extreme lower (100 to 800")) %>%
  filter(!is.na(value))
    
#table(geo_long$group)

geo_plot <- ggplot(geo_long, aes(group)) + 
  geom_bar()+ 
  facet_wrap(~ dem_type) + 
  coord_flip()

geo_plot 

```

# spatial uncertainty 

```{r}
geo_longsf <- st_sf(geo_long)

geo_longsf %>%
  ggplot() + 
  geom_sf(aes(colour = dem_type)) + 
  facet_wrap(~ group)
